%
% METU Institute of Natural and Applied Sciences Thesis example
%
% Edited and Commented by Utku Erdoğdu 2012
%
% Please read the explanations so that you can customize the document
%
% Files needed by this document:
% metu.cls
% metu11.def (if you will use 11pt fonts)
% metu12.def (if you will use 12pt fonts)
% metu10.def (if you will use 10pt fonts)
%
% Possible Options Here:
%
% oneandhalf, double, single : Line spacing used in the thesis. Default and institute preference is
% single.
%
% 10pt, 11pt, 12pt : Font size Default is 10pt, which is institue choice.
%
% pntr, pntc, pnbt : Page number position. Options are top center, top right or bottom. Default and
% institute preference is page numbers at bottom. When page numbers are at the top bottom margins
% are skewed.
%
% chaproman, chaparabic: Chapter numbering format. Options are roman numbers and arabic numbers.
% Default is roman, institute prefers arabic
%
% oneside, twoside : Printing style. Default is twoside, which is institute choice. In this style
% chapters begin from odd numbered pages.
%
% tr, eng : Document language. This is useful if you want to translate your thesis into
% Turkish. Then you give the option tr and use \ifturkish. . .\else. . .\fi  whenever you want
% to do something only for Turkish or only for English. Default is eng.
% IMPORTANT!! : For official institute documents you should not use this option.
% The Turkish format is only supplied for custom translations.
%
% ceng,aee,arme.. : You can use the abbreviated form of your department here and there is no further need to
% define the department name below. If your department name is not among the below list of defined
% departments, you should use \department and \turkishdepartment macros to define the name of your
% department.
%
% Defined Departments and Abbreviations:
% --------------------------------------
% Computer Engineering : ceng
% Aerospace Engineering : aee
% Archaeometry : arme
% Architecture : arch
% Biochemistry : bch
% Biology : biol
% Biomedical Engineering : bme
% Biotechnology : btec
% Building Science : bs
% Cement Engineering : ceme
% Chemical Engineering : che
% Chemistry : chem
% City and Regional Planning : crp
% City Planning : cp
% Civil Engineering : ce
% Computational Design and Fabrication Technologies in Architecture : arcd
% Computer Education and Instructional Technology : cte
% Design Research for Interaction : iddi
% Earthquake Studies : eqs
% Earth System Science : ess
% Electrical and Electronics Engineering : ee
% Engineering Management : em
% Engineering Sciences : es
% Environmental Engineering : enve
% Food Engineering : fde
% Geodetics - Geographical Information Technologies : ggit
% Geological Engineering : geoe
% Hydrosystems Engineering : he
% Industrial Design : id
% Industrial Engineering : ie
% Mathematics : math
% Mechanical Engineering : mech
% Metallurgical and Materials Engineering : mete
% Micro and Nanotechnology : mnt
% Mining Engineering : mine
% Operational Research : or
% Petroleum and Natural Gas Engineering : pete
% Physics : phys
% Polymer Science and Technology : pst
% Regional Planning : rp
% Restoration : rest
% Secondary Science and Mathematics Education : ssme
% Software Engineering : se
% Statistics : stat
% Structural Mechanics : st
%
% phd, ms : Degree Received. Ph.D. or M.S. Default is M.S.
%
% End of Options
\documentclass[11pt,single,chaparabic,ceng,ms,eng]{metu}
% You can delete next line If your thesis does not have an appendix
\usepackage{appendix}

%
% Use your latex packages here
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{colortbl}
\usepackage{array}
\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{float}
\usepackage{subfig}
\usepackage{rotating}
\usepackage{tikz}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{slashbox}
\usepackage{verbatim}
\usepackage{hhline}
\definecolor{Gray}{gray}{0.85}
\definecolor{LightCyan}{rgb}{0.88,1,1}
% End of Latex Packages
%
% Any personal Latex definition, decleration, etc.


% End of personal stuff
%
% Personal Information
% ----------------------------
%
% Please check this part and fill in information about your thesis
%
% Name and Surname
\author{Saber HafezQorani}
% Thesis Title English and Turkish
\title{MODELING THE COMBINED EFFECT OF RNA-BINDING PROTEINS AND MICRORNAS IN POST-TRANSCRIPTIONAL REGULATION}
\turkishtitle{TRANSKRIPSIYON-SONRASI KONTROLDE RNA'YA BAGLANAN PROTEIN'LER VE MIKRORNA'LARIN ETKİSİNİN BİRLİKTE MODELLENMESİ}
% Department : English and Turkish
%
% Some of the departments are pre-defined, you need not redeclare them. You can use them by just
% giving an option to \documentclass. See documentation for options above. If you will define your
% department here do not use ``Department'' or ``Bölümü'' words.
%\department{Computer Engineering}
%\turkishdepartment{Bilgisayar Mühendisliği}
%
%
% Date : You should indicate the month of your thesis defence in English.
% Default is this month
%
\date{September 2015}
%
% Approval Page Details
% --------------------------
% For each command you can give the title as optional parameter enclosed in [ ]
% This also handles the Turkish titles if you're planning to produce Turkish version of the
% document. If you'll hard code the title, you need to use turkish version of each command after
% the command itself
%
% prof : Prof. Dr.
% assocprof : Assoc. Prof. Dr.
% assistprof : Assist. Prof. Dr.
% dr : Dr.
%
% Director of Institute
\director[prof]{Nazife Baykal}
% Head of Department
\headofdept[assocprof]{Yeşim Aydın Son}
%
% Supervisor : English and Turkish
\supervisor[assocprof]{Yeşim Aydın Son}
% \turkishsupervisor{  } %if you will hard-code the academic title
%
% Affiliation of Supervisor in English and possibly in Turkish
\departmentofsupervisor{Health Informatics, METU}
% Co Supervisor if Any : English and Turkish
% You can just delete the next lines if you don't have a co-supervisor
\cosupervisor[assistprof]{Hilal Kazan}
% \turkishcosupervisor{Prof. Dr. Reda Alhajj} %if you will hard-code the academic title
% Affiliation of Co-Supervisor
% You can just delete the next line if you don't have a co-supervisor
\departmentofcosupervisor{Computer Engineering, Antalya International University}
%
% Committee Members
% In general members are sorted according to their academic titles
%
% Proffesors (1)
% Associate Professors (2)
% Assistant Professors (3)
% Other (4)
%
% IMPORTANT:  All affiliatons should fit in a single line
% If affiliation line is broken into two lines you should shorten the affiliation by using
% abbrevations or any other means
%
% First committee member should be the chair of examining committee
% Typically the chair is one of the highest ranked committee members
% Ask your supervisor if you are not sure
\committeememberi[]{Aybar Can Acar}
% \turkishcommitteememberi{Prof. Varol Akman} % if you will hard-code the academic title
\affiliationi{Health Informatics, METU}
% Second committee member is always your supervisor
\committeememberii{Yeşim Aydın Son}
\affiliationii{Health Informatics, METU}
% If you are an M.Sc. student and your Co-Supervisor is in your
% examination committee, then third committee member is always your co-supervisor
%
% IMPORTANT: If you are Ph.D. student your co-supervisor can not be in your
% examination committee.
\committeememberiii{Hilal Kazan}
\affiliationiii{Computer Engineering, AIU}
% Fourth committee member
\committeememberiv{Tolga Can}
\affiliationiv{Computer Engineering, METU}
% Fifth committee member
\committeememberv{Bala Gür Dedeoğlu}
\affiliationv{Biotechnology Institute, Ankara University}
%
% Keywords : English & Turkish, Comma seperated
\keywords{Post-transcriptional regulation, RNA-binding proteins, MicroRNAs, RNA secondary structure, mRNA stability}
\anahtarklm{Transkripsiyon-sonrası kontrol, RNA'ya bağlanan protein, MikroRNA, RNA'ın ikincil yapısı, gen stabilitesi}
%
% Abstract in English
%
\abstract{Post-transcriptional regulation (PTR) controls the gene expression between transcription and translation. Regulation at this level is carried out by the interactions of trans-acting RNA-binding proteins (RBPs) and microRNAs (miRNAs) with cis-regulatory elements in mRNA. Majority of previous work have focused on the effect of a single factor independent of other co-factors bound to the same mRNA. However, recent studies have shown that RBPs and miRNAs can act in cooperation or competition with each other. In this thesis, we mapped the binding sites of both RBPs and miRNAs on human 3’UTRs, and utilized this collection of binding sites to better understand PTR networks. We first focused on several RBPs and assessed how accessibility and conservation differ between experimentally supported sites and other sites that are only computationally predicted. We then investigated the competitive effects of other factors on HuR binding and the resulting transcript abundance change upon HuR depletion. Next, we characterized the potential interactions between the factors by finding those pairs of factors with co-occurrence of motifs higher than expected by chance. Our results show that PUM1 and PUM2 have potential cooperative interactions with miRNAs. Finally, we used logistic regression with features compiled from the counts of sites of factors and dinucleotide frequency to accurately predict the stability and steady-state abundance of mRNAs. Altogether, results of this thesis suggest that studies of PTR must consider the effect of both RBPs and miRNAs, and their interactions.}
%
% Turkish Abstract
%
\oz{Transkripsiyon-sonrası kontrol (TSK), gen ifadesinin transkripsiyon ile translasyon arasındaki adımlarını kontrol eder. Bu adımların kontrolü RNA'ya bağlanan protein (RBP) ve mikroRNA'ların (miRNA) mesajcı RNA'lardaki (mRNA) hedef noktalarına bağlanmaları sayesinde gerçekleştirilir. Şu ana kadar yapılan çalışmaların büyük çoğunluğunda tek bir RBP ya da tek bir miRNA aynı mRNA'ya bağlanan diğer faktörlerden bağımsız olarak incelenmiştir. Ancak, son çalışmalar RBP ve miRNA'ların birbirleriyle işbirliği ya da rekabet ilişkileri içerisinde olduklarını göstermiştir. Bu tezde, son gelişmelere paralel olarak insan 3'UTR'ları üzerindeki hem RBP, hem de miRNA bağlanma noktalarını belirledik. İlk olarak, üzerinde çalışılmış RBP'lerin deneysel yöntemlerle bulunan bağlanma noktaları ile işlemsel yöntemlerle tahmin edilmiş bağlanma noktalarını RNA'nın ikincil yapısı ve evrimsel korunum bakımından farklı olduğunu gösterdik. Daha sonra, HuR adlı RBP için var olan susturulma sonrası mRNA ifade değişimi verilerini kullanarak diğer faktörlerin HuR'la olan rekabetçi ilişkilerini araştırdık. Ayrıca, iki faktörün (RBP ya da miRNA) bağlanma noktalarının beraber görülme sıklığını hesaplayarak potansiyel işbirliği ilişkilerini inceledik. Bu analiz sonucunda PUM1 ve PUM2 RBP'lerinin miRNA'larla işbirliği içerisinde olduğunu gözlemledik. Son olarak, faktör bağlanma sayıları ve dinükleotid frekansı gibi öznitelikleri bağlanım modeliyle kullanarak gen ifadesi ve stabilitesini yüksek bir doğruluk payıyla tahmin ettik. Elde ettiğimiz sonuçlar transkripsiyon-sonrası kontrol ile ilgili mekanizmaları daha iyi anlamak için RBP'lerin, miRNA'ların ve ayrıca aralarındaki ilişkilerin de göz önünde bulundurulması gerektiğine işaret etmektedir.}
%
% Dedication
\dedication{Dedicated to my family and wife.}
%
%
% Acknowledgements
\acknowledgments{
I would like to express my gratitude to my co-supervisor Assist. Prof. Dr. Hilal Kazan whose expertise, understanding, and patience, added considerably to my graduate experience. It was a great honor to work with her for the last two years and our cooperation influenced my academical view in a great scale. I also would like to thank Assoc. Prof. Dr. Yeşim Aydın Son for her support of my thesis work.

This project is supported by TÜBİTAK (Project No: #113E159). I worked on this project along with my colleague, Atefeh Lafzi, under supervision of Assist. Prof. Dr. Hilal Kazan for the previous two years. Our work also accepted as poster and short communication at several conferences including ISMB 2014 (22nd Annual International Conference on Intelligent Systems and Molecular Biology), MIE 2014 (The 25th International Congress of the European Federation for Medical Informatics), and ECCB 2014 (13th European Conference on Computational Biology). We also submitted our work as a paper to the Nucleic Acid Research Journal and it is under review. I would like to thank Atefeh Lafzi who supported and helped me in this project.

I also had the chance to continue my work on this project at Antalya International University. I would like to thank all the faculty, administrative staff and research assistants at AIU who made me feel welcome there.

I would also like to thank my family for the support they provided me through my entire life and in particular, I specially thank my wife, Mahsa Ghavidel, for her endless support, love and encouragement. I am also thankful for my parents, Esmaiel HafezQorani and Zahra Rafie for their great support during every aspect of my life. I would like to specially thank my older brother, Saleh HafezQorani, who always kindly guided me in my life and supported my academic career.
}
%

% End of Personal and Introductory Information
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5

%%% !!! This two should be last lines before \begin{document}, do no move them !!!
\usepackage[pdftex]{hyperref}
\usepackage[all]{hypcap}
\begin{document}
% Preliminaries
\begin{preliminaries}
% If you are willing to use any custom stuff before Chapters, put it here
% Such as List of Abbreviations
% Check the abbreviations.tex for a template list of abbreviations
% End of Preliminaries
\end{preliminaries}
%
% Latex content Goes Here
%
%
\input{ch1_introduction/chapter1.tex}
\input{ch2_background/chapter2.tex}
\input{ch3_materials_methods/chapter3.tex}
\input{ch4_results_discussion/chapter4.tex}
\input{ch5_conclusion_futurework/chapter5.tex}
%
% References in Bibtex format goes into below indicated file with .bib extension
\bibliography{referanslar}
\bibliographystyle{unsrt}
% You can use full name of authors, however most likely some of the Bibtex entries you will find, will use abbreviated first names
% If you don't want to correct each of them by hand, you can use abbreviated style for all of the references
\appendix
\input{appendix1/appendix1.tex}
\input{appendix2/appendix2.tex}
%
% If you are a Ph.D. Student you need to insert a CV at the end of you thesis
% Check vita.tex for a simple CV template in Latex

\end{document}
